Spring haul-out behavior of seals in the Bering and Chukchi seas

### 2023-02-27 analysis/paper/London_HauloutBeahvior.Rmd
additional responses to reviews
quite a bit in this commit. really focused on rearranging the location for some of the text, adding deployment location figure, and expanding the discussin re: not including sea ice as a covariate.

  - \usepackage {newtxtext,newtxmath}

@@

  - \usepackage{setspace}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}

@@

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  include = FALSE,
  fig.path = "../figures/",
  dpi = 300,
  out.width = "100%"
)

@@

Statutory requirements (e.g., United States Marine Mammal Protection Act (MMPA),
United States Endangered Species Act (ESA)) for timely estimates of population
abundance and trend of these species mean improved aerial survey effort is
needed -- and, must be paired with improved knowledge of haul-out behavior to
ensure appropriate survey design, robust methods, and accurate estimates.
Several studies have contributed estimates of the distribution and abundance of
ice-associated seal species in the Arctic using aerial surveys (e.g.,
[@bengtson2005a], [@conn2014a], and [@verhoef2014a]). Such abundance studies
were conducted over very large areas and estimation of absolute abundance
required making inference about numerous issues affecting the observation of
seals on ice. These included availability (only seals on ice were available to
be counted), detection probability (observers or automated detection systems may
have missed some seals on ice), species misclassification, and possible
disturbance of seals by aircraft [@conn2014a; @verhoef2014a]. Refining these
inferences will improve the accuracy of abundance estimates and, hopefully,
allow credible predictions about the effects of climate disruptions on the
abundance and distribution of Arctic seal populations.

@@

For this study we used haul-out behavior data and location estimates from
bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
Chukchi, and western Beaufort seas by multiple organizations as part of
collaborative investigations from 2005 through 2020. Seals were captured using
nets and bio-loggers were attached during studies based in coastal Alaska
communities or on research ships. Ship-based capture events occurred during
spring near the southern ice edge in the Bering Sea between 2005 and 2018.
Land-based capture events occurred between 2005 and 2020 from May to October,
generally between the coastal communities of Scammon Bay, Alaska in the Bering
Sea, Utqiaġvik, Alaska in the Chukchi Sea, and Nuiqsut in the Beaufort Sea. We
refer readers to the primary literature for detailed capture and bio-logger
attachment methods (see Supplemental Material, S1).

@@

deploy_table %>%
  dplyr::group_by(species, sex, age) %>%
  dplyr::summarise(
    n = n(),
    seal_hours = sum(n_hours)
  ) %>%
  dplyr::mutate(
    deploy_stats = glue::glue('{n} ({format(seal_hours, big.mark=",")} seal hours)')
  ) %>%
  dplyr::select(-c("n","seal_hours")) %>%
  tidyr::pivot_wider(names_from = age, values_from = c(deploy_stats)) %>%
  dplyr::ungroup() %>%
  dplyr::rename(Species = species, Sex = sex, `Young-of-the-Year` = `YOUNG OF YEAR`,
                `Subadult` = SUBADULT, Adult = ADULT) %>%
  flextable() %>%
  add_header(`Young-of-the-Year` = "Age Class",
   Subadult = "Age Class", Adult = "Age Class",
   top = TRUE ) %>%
  merge_h(part = "header") %>%
  align(align = "left", part = "header") %>% 
  align(i = 1, align = "center", part = "header") %>% 
  fontsize(size = 10, part = "all") %>% 
  width(width = 1.35) %>% 
  width(j = ~ Species, width = 0.9) %>% 
  width(j = ~ Sex, width = 0.4)

@@

Of key interest in this study was the relationship between haul-out behavior and
weather covariates that vary with time and seal location. We explored the use
of a continuous-time correlated random walk [@johnson2008a] movement model
to predict locations at specific times. However, the sparse nature of data from
some bio-loggers, especially those mounted to the rear flipper, resulted in poor
modeling performance or convergence issues. For this study, we calculated a
weighted average daily location where the inverse of the estimated Argos or
FastLoc GPS location error was used for the weight. Each Argos location estimate
was assigned an error radius based on either the categorical location quality (
_3_ = 250 m, _2_ = 500 m, _1_ = 1500 m, _0_ = 2500 m [@lopez2013]; we chose 2500 m for
location classes _A_ and _B_) or, when available, the estimated error radius
from the Argos Kalman filter algorithm. Location estimates from FastLoc GPS were
all assigned an error radius of 50 m. On days when haul-out observations were
present but location data were missing we used the seal's last calculated weighted
average daily location; days when the location intersected with land
were removed from the seal's record. We recognize that bearded and spotted
seals haul out on land. However, assessing the relationship between haul-out
behavior and weather covariates and seals' availability for aerial surveys on
land was outside the scope of this study. Additionally, any daily locations on
land were likely more reflective of coordinate averaging and measurement error,
rather than actual use of coastal features. Figure \@ref(fig:dataMap) shows the
spatial distribution of weighted locations with available haul-out behavior data
used for analysis of each species across the study area.

```{r dataMap, fig.height = 3.75, include=TRUE, fig.cap="(ref:mapCaption)"}

ggplot() +
  annotation_spatial(world, 
                     size = 0, 
                     fill = "grey70") +
  layer_spatial(
    data = locs_count, size = 0.125,
    aes(fill = ct)
  ) +
  scale_fill_gradientn(colors = rev(met.brewer("Greek")),
                    trans = "log10", aesthetics = "fill",
    name = "number of locations",
    guide = guide_colorbar(title.position = "bottom", barwidth = 15, 
                           barheight = 0.5, title.hjust = 0.5)) + 
  facet_grid(~ species) +
  theme(legend.position = "bottom") +
  ggtitle("Spatial distribution of haul-out behavior data") +
  labs(caption = bquote("One hexagonal cell represents 50"~km^2))
```

@@

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem
counterintuitive when modeling the haul-out behavior of seal species with such a
close association to sea ice; seals haul out in the presence of sea ice, and we
could assess the local concentration of sea ice during these events (see
[@olnes2020]). This, however, expands the scope of our analysis into the realm
of habitat selection and many of our deployments consisted of a single device
attached to the rear flipper of the seal which did not provide at-sea locations,
limiting our ability to fully evaluate fine-scale habitat preferences related to
sea-ice. Insight into how seals use and interact with sea-ice during an extended
period when the availability and characteristics of sea-ice is rapidly changing
is important but ancillary to the focus of this analysis. The focus of this
study was to develop models applicable for aerial survey correction factors and
using sea ice as a covariate would almost certainly bias haul-out predictions
towards those seals that are on or near ice and therefore more likely to haul
out. Since aerial surveys can only detect seals on ice, abundance estimates
would be missing a correction for those seals that are away from ice. Lastly,
our study was limited to the spring season when seal haul-out behaviors are
strongly influenced by pupping, nursing, breeding behavior, and molt and these
drivers are likely more influential than specific sea-ice concentration.
Crawford et al. [-@crawford2019] compared haul-out probability models for ringed
seals and found those that only included season (and not sea-ice concentration)
were the most parsimonious. For these reasons, we have elected not to use sea
ice concentration _as a predictor for haul-out probability_ in the present
study.

@@

```{r tblCovariates, tab.id = "tblCovariates", include=TRUE, tab.cap = "Explanatory covariates used in analyses of binary haul-out records for bearded, spotted, and ribbon seals. Note that we also considered select interactions (see article text) between these primary covariates.  For instance, wind chill was represented by the interaction _temperature:wind_."}

@@

covariate_tbl %>% 
  select(-Source) %>% 
  flextable() %>% 
  fontsize(size = 10, part = "all") %>% 
  hline(i=8) %>% 
  width(width = 1.0) %>% 
  width(j = ~ Description, width = 2.5)
```

@@

Figure \@ref(fig:dataCal) shows the distribution of all
haul-out data across the study season for each species. Observations for ribbon
and spotted seals were concentrated in the months of May and June due to the
timing of deployment (April and May) and the timing of molt (May and June).
During molt, seals (and their attached bio-loggers) spend more time out of the
water and more data are transmitted. Molt timing also impacts when many
deployments end as any bio-loggers adhered to the hair will fall off. Relative
to the other species in the study, there were fewer deployments of bio-loggers
on bearded seals. This resulted in fewer data observations overall and
noticeably lower in numbers May and June. The majority of bearded seal
deployments started later in the summer and, by May, bio-loggers had either
fallen off or their batteries were depleted.

```{r dataCal, include=TRUE, fig.height = 4, fig.cap = "Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species. Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020. The higher density of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, many bio-logger deployments started in April and May. The overall reduced quantity of observations from bearded seals is reflective of the lower number of bio-logger deployments in the study."}

ggplot(data = seal_hours_tbl, aes(x = yday, y = n, fill = n))+
  geom_col() +
        scale_fill_gradientn(colors = met.brewer("VanGogh3")) +
        # scale_fill_viridis_c(option="mako") +
        scale_x_continuous(breaks = c(60, 91, 121, 152, 182),
                           labels = c("01 March", "01 April", "01 May", "01 June", "01 July"),
                           expand = expansion(mult = c(0, 0))) +
        scale_y_continuous(expand = expansion(mult = c(0, 0))) +
  guides(fill = guide_colorbar(title.position = "bottom", 
                               title.hjust = 0.5, barwidth = 15,
                               barheight = 0.5)) +
  facet_grid(species ~ .) + 
  theme(axis.text.x = element_text(hjust = 0),
        axis.text.y = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(fill = "seal-hours of bio-logger data") +
  ggtitle("Density of haul-out behavior observations varies from March through July",
          subtitle = "data are grouped by day-of-year and presented in seal-hours collated across years (2005-2020)")

@@

```{r beardedHOCal, fig.height = 4.5, include=TRUE, fig.cap="Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes. Hour of the day was fixed at local solar noon."}

@@

```{r beardedHOwx, fig.height = 5.5, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications. Hour of the day was fixed at local solar noon and day-of-year held at the day of peak haul-out for bearded seals. Vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}

@@

```{r ribbonHOCal, fig.height = 5.5, include=TRUE, fig.cap="Predicted hourly haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model. Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males. 'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes. Hour of the day was fixed at local solar noon."}

@@

```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of ribbon seals within each age and sex classification. Hour of the day was fixed at local solar noon and day-of-year held at the average day of peak haul-out for ribbon seals. Vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}

@@

```{r spottedHOCal, fig.height = 5.5, include=TRUE, fig.cap="Predicted hourly haul-out probability of spotted seals from 1 March to 15 July for each age and sex class used in the model. Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males. 'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes. Hour of the day was fixed at local solar noon."}

@@

```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of spotted seals within each age and sex classification. Hour of the day was fixed at local solar noon and day-of-year held at the average day of peak haul-out for spotted seals. Vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}

@@

```{r annualHO, fig.height = 5.5, include=TRUE, fig.cap="Annual variability in the timing of peak haul-out probability (colored markers) for ribbon and spotted seals across 13 years. Predictions are shown for local solar noon and under smoothed weather conditions. Only those groups (age:sex + year) that included observations from more than one seal are shown. Additionally, any groups where data were only available after 1 June or before 1 May are not included."}

@@

>>>>

  - \usepackage{stix2}
  - \usepackage {booktabs}
  - \usepackage {setspace}

@@

  - \usepackage {pdflscape}
  - \newcommand {\blandscape}{\begin{landscape}}
  - \newcommand {\elandscape}{\end{landscape}}
  - \usepackage {caption}
  - \captionsetup{font={normalsize,onehalfspacing}}

@@

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  include = FALSE,
  fig.path = "../figures/",
  dpi = 300,
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "100%",
  tab.topcaption = TRUE
)

@@

Statutory requirements (e.g., United States Marine Mammal Protection Act (MMPA),
United States Endangered Species Act (ESA)) for timely estimates of population
abundance and trend of these species mean improved aerial survey effort is
needed -- and, must be paired with improved knowledge of haul-out behavior to
ensure appropriate survey design, robust methods, and accurate estimates.
Several studies have contributed estimates of the distribution and abundance of
ice-associated seal species in the Arctic using aerial surveys (e.g.,
[@bengtson2005a], [@conn2014a], and [@verhoef2014a]) and more recent efforts
have significantly expanded on previous survey effort. Such abundance studies
are conducted over very large areas and estimation of absolute abundance
requires making inference about numerous issues affecting the observation of
seals on ice. These include availability (only seals on ice are available to
be counted), detection probability (observers or automated detection systems may
miss some seals on ice), species misclassification, and possible
disturbance of seals by aircraft [@conn2014a; @verhoef2014a]. Refining these
inferences will improve the accuracy of abundance estimates and, hopefully,
allow credible predictions about the effects of climate disruptions on the
abundance and distribution of Arctic seal populations.

@@

For this study we used haul-out behavior data and location estimates from
bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering,
Chukchi, and western Beaufort seas by multiple organizations as part of
collaborative investigations from 2005 through 2020. Seals were captured using
nets and bio-loggers were attached during studies based in coastal communities
or on research ships. Ship-based capture events occurred during spring near the
southern ice edge in the Bering Sea between 2005 and 2018. Land-based capture
events occurred between 2005 and 2020 from May to October, generally between the
coastal communities of Scammon Bay, Alaska in the Bering Sea, Utqiaġvik, Alaska
in the Chukchi Sea, and Nuiqsut in the Beaufort Sea. Data from additional
deployments along the Kamchatka peninsula in the western Bering Sea are also
included. We refer readers to Figure \@ref(fig:deployplot) and the primary literature
for detailed capture and bio-logger attachment methods (see Supplemental
Material, S1).

(ref:deployMapCaption) Initial deployment areas for each seal species (bearded, ribbon, and spotted seals) in the study between 2005 and 2020. Deployments were initiated across a range of months but only data from 1 March to 15 July were included in the analysis. See Figure \@ref(fig:dataMap) for the spatial distribution of observed data used in the study.

```{r deployplot, include=TRUE, fig.asp = 0.45, fig.cap = "(ref:deployMapCaption)"}
targets::tar_load(deploy_details, store = here::here("_targets"))

deploy_sf <- deploy_details %>% 
  sf::st_as_sf(coords = c('longitude','latitude'), crs = 4326) %>% 
  sf::st_transform(3571) %>% 
  dplyr::group_by(species,location, region) %>% 
  dplyr::group_modify(~ concaveman::concaveman(., concavity = 4),.keep=TRUE) %>% 
  sf::st_sf(crs = 3571) %>% 
  sf::st_buffer(40*1000)

world <- ne_countries(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(st_crs(deploy_sf))

species_colors <- setNames(met.brewer("Isfahan2",n=3,direction=-1,override.order = TRUE),
                           c("Bearded seal","Ribbon seal","Spotted seal"))
species_labels <- c("Bearded seal" = paste0("<span style = 'color:",
                                            species_colors['Bearded seal'],
                                            ";'>Bearded seal</span>"),
                    "Ribbon seal" = paste0("<span style = 'color:",
                                           species_colors['Ribbon seal'],
                                           ";'>Ribbon seal</span>"),
                    "Spotted seal" = paste0("<span style = 'color:",
                                            species_colors['Spotted seal'],
                                            ";'>Spotted seal</span>")
) %>% as_labeller()

p <- ggplot() +
  layer_spatial(deploy_sf, aes(fill=species),
                linewidth=0) +
  annotation_spatial(world,
                     linewidth=0,
                     fill = "grey80") +
  facet_grid(cols = vars(species), labeller = species_labels,
             switch="x") +
  scale_fill_manual(values = met.brewer("Isfahan2",n=3,direction=-1,override.order = TRUE),
                    aesthetics = "fill", 
                    name = "species") + 
  ggtitle("Areas of initial deployment location",
          subtitle = 
                 paste0("Bio-loggers were initially deployed on ",
                "<span style = 'color:",species_colors['Bearded seal'],";'>bearded, </span> ", 
                "<span style = 'color:",species_colors['Ribbon seal'],";'>ribbon, </span> ",  
                "and ", 
                "<span style = 'color:",species_colors['Spotted seal'],";'>spotted seals</span> ", 
                "across multiple regions within the Bering and Chukchi seas.")) +
  theme(legend.position = "none",
        strip.background = element_blank(),
        strip.placement = "outside",
        strip.text.x = element_markdown(face="bold"),
        plot.subtitle = element_textbox_simple(margin = ggplot2::margin(b=12)))

p
```

@@

deploy_table %>%
  dplyr::group_by(species, sex, age) %>%
  dplyr::summarise(
    n = n(),
    seal_hours = sum(n_hours)
  ) %>%
  dplyr::mutate(
    deploy_stats = glue::glue('{n} ({format(seal_hours, big.mark=",")} seal hours)')
  ) %>%
  dplyr::select(-c("n","seal_hours")) %>%
  tidyr::pivot_wider(names_from = age, values_from = c(deploy_stats)) %>%
  dplyr::ungroup() %>%
  dplyr::rename(Species = species, Sex = sex, `Young-of-the-Year` = `YOUNG OF YEAR`,
                `Subadult` = SUBADULT, Adult = ADULT) %>%
  flextable() %>%
  add_header(`Young-of-the-Year` = "Age Class",
   Subadult = "Age Class", Adult = "Age Class",
   top = TRUE ) %>%
  merge_h(part = "header") %>%
  align(align = "left", part = "header") %>% 
  align(i = 1, align = "center", part = "header") %>% 
  fontsize(size = 9, part = "all") %>% 
  width(width = 1.35) %>% 
  width(j = ~ Species, width = 0.9) %>% 
  width(j = ~ Sex, width = 0.4)

@@

Of key interest in this study was the relationship between haul-out behavior and
weather covariates that vary with time and seal location. We explored the use
of a continuous-time correlated random walk [@johnson2008a] movement model
to predict locations at specific times. However, the sparse nature of data from
some bio-loggers, especially those mounted to the rear flipper, resulted in poor
modeling performance or convergence issues. For this study, we calculated a
weighted average daily location where the inverse of the estimated Argos or
FastLoc GPS location error was used for the weight. Each Argos location estimate
was assigned an error radius based on either the categorical location quality (
_3_ = 250 m, _2_ = 500 m, _1_ = 1500 m, _0_ = 2500 m [@lopez2013]; we chose 2500 m for
location classes _A_ and _B_) or, when available, the estimated error radius
from the Argos Kalman filter algorithm. Location estimates from FastLoc GPS were
all assigned an error radius of 50 m. On days when haul-out observations were
present but location data were missing we used the seal's last calculated weighted
average daily location; days when the location intersected with land
were removed from the seal's record. We recognize that bearded and spotted
seals haul out on land. However, assessing the relationship between haul-out
behavior and weather covariates and seals' availability for aerial surveys on
land was outside the scope of this study. Additionally, any daily locations on
land were likely more reflective of coordinate averaging and measurement error,
rather than actual use of coastal features.

@@

@@

```{r tblCovariates, include=TRUE, tab.cap = "Explanatory covariates used in analyses of binary haul-out records for bearded, spotted, and ribbon seals. Note that we also considered select interactions (see article text) between these primary covariates.  For instance, wind chill was represented by the interaction temperature:wind."}

@@

covariate_tbl %>% 
  select(-Source) %>% 
  flextable() %>% 
  fontsize(size = 9, part = "all") %>% 
  hline(i=8) %>% 
  width(width = 1.0) %>% 
  width(j = ~ Description, width = 2.5)
```

@@

Figure \@ref(fig:dataMap) shows the
spatial distribution of weighted locations with available haul-out behavior data
used for analysis of each species across the study area.

```{r dataMap, include=TRUE, fig.cap="(ref:mapCaption)"}

ggplot() +
  annotation_spatial(world, 
                     size = 0, 
                     fill = "grey80") +
  layer_spatial(
    data = locs_count, size = 0.125,
    aes(fill = ct)
  ) +
  scale_fill_gradientn(colors = rev(met.brewer("Greek")),
                    trans = "log10", aesthetics = "fill",
    name = "number of locations",
    guide = guide_colorbar(title.position = "bottom", barwidth = 15, 
                           barheight = 0.5, title.hjust = 0.5)) + 
  facet_grid(~ species) +
  theme(legend.position = "bottom") +
  ggtitle("Spatial distribution of haul-out behavior data") +
  labs(caption = bquote("One hexagonal cell represents 50"~km^2))
```

Figure \@ref(fig:dataCal) shows the temporal distribution of all
haul-out data across the study season for each species. Observations for ribbon
and spotted seals were concentrated in the months of May and June due to the
timing of deployment (April and May) and the timing of molt (May and June).
During molt, seals (and their attached bio-loggers) spend more time out of the
water and more data are transmitted. Molt timing also impacts when many
deployments end as any bio-loggers adhered to the hair will fall off. Relative
to the other species in the study, there were fewer deployments of bio-loggers
on bearded seals. This resulted in fewer data observations overall and
noticeably lower in numbers May and June. The majority of bearded seal
deployments started later in the summer and, by May, bio-loggers had either
fallen off or their batteries were depleted.

```{r dataCal, include=TRUE, fig.cap = "Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species. Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020. The higher density of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, many bio-logger deployments started in April and May. The overall reduced quantity of observations from bearded seals is reflective of the lower number of bio-logger deployments in the study."}

ggplot(data = seal_hours_tbl, aes(x = yday, y = n, fill = n))+
  geom_col() +
        scale_fill_gradientn(colors = rev(met.brewer("Greek"))) +
        scale_x_continuous(breaks = c(60, 91, 121, 152, 182),
                           labels = c("01 March", "01 April", "01 May", "01 June", "01 July"),
                           expand = expansion(mult = c(0, 0))) +
        scale_y_continuous(expand = expansion(mult = c(0, 0))) +
  guides(fill = guide_colorbar(title.position = "bottom", 
                               title.hjust = 0.5, barwidth = 15,
                               barheight = 0.5)) +
  facet_grid(species ~ .) + 
  theme(axis.text.x = element_text(hjust = 0),
        axis.text.y = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(fill = "seal-hours of bio-logger data") +
  ggtitle("Density of haul-out behavior observations varies from March through July",
          subtitle = "data are grouped by day-of-year and presented in seal-hours collated across years (2005-2020)")

@@

```{r beardedHOCal, include=TRUE, fig.cap="Predicted hourly haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes. Hour of the day was fixed at local solar noon."}

@@

```{r beardedHOwx, fig.asp = 1, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications. Hour of the day was fixed at local solar noon and day-of-year held at the day of peak haul-out for bearded seals. Vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}

@@

```{r ribbonHOCal, fig.asp = 0.9, include=TRUE, fig.cap="Predicted hourly haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model. Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males. 'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes. Hour of the day was fixed at local solar noon."}

@@

```{r ribbonHOwx, fig.asp = 0.9, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of ribbon seals within each age and sex classification. Hour of the day was fixed at local solar noon and day-of-year held at the average day of peak haul-out for ribbon seals. Vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}

@@

```{r spottedHOCal, fig.asp = 0.9, include=TRUE, fig.cap="Predicted hourly haul-out probability of spotted seals from 1 March to 15 July for each age and sex class used in the model. Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males. 'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes. Hour of the day was fixed at local solar noon."}

@@

```{r spottedHOwx,fig.asp = 0.9, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and precipitation on the predicted haul-out probability of spotted seals within each age and sex classification. Hour of the day was fixed at local solar noon and day-of-year held at the average day of peak haul-out for spotted seals. Vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}

@@

```{r annualHO, fig.asp = 0.75, include=TRUE, fig.cap="Annual variability in the timing of peak haul-out probability (colored markers) for ribbon and spotted seals across 13 years. Predictions are shown for local solar noon and under smoothed weather conditions. Only those groups (age:sex + year) that included observations from more than one seal are shown. Additionally, any groups where data were only available after 1 June or before 1 May are not included."}

@@

Notably missing from our list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem
counterintuitive when modeling the haul-out behavior of seal species with such a
close association to sea ice; seals haul out in the presence of sea ice, and we
could assess the local concentration of sea ice during these events (see
[@olnes2020]). This, however, expands the scope of our analysis into the realm
of habitat selection and many of our deployments consisted of a single device
attached to the rear flipper of the seal which did not provide at-sea locations,
limiting our ability to fully evaluate fine-scale habitat preferences related to
sea-ice. Insight into how seals use and interact with sea-ice during an extended
period when the availability and characteristics of sea-ice is rapidly changing
is important but ancillary to the focus of this analysis. The focus of this
study was to develop models applicable for aerial survey correction factors and
using sea ice as a covariate would almost certainly bias haul-out predictions
towards those seals that are on or near ice and therefore more likely to haul
out. Since aerial surveys can only detect seals on ice, abundance estimates
would be missing a correction for those seals that are away from ice. Consider
the case of employing haul-out probabilities as an aerial survey correction in a
population inhabiting two areas: an ice-free region and a region with sea ice.
Further, assume that half of the population is in each area on average, and that
the probability of hauling out is 1.0 in the ice covered region and 0.0 in the
region without ice.  Denoting the total population as $N$, let us examine what
happens when we use (1) population level availability, and (2) ice-specific
availability as a correction factor.  We'll further assume, for sake of this
example, that detection probability of seals on ice is 1.0, and that the
entirety of the ice covered region is censused. In case (1), our aerial survey
count is $C = 0.5 \cdot N$, with a population-level availability probability of
$\hat{a} = 0.5$. A Horvitz-Thompson-type estimator for abundance is simply 
$\hat{N} = C \div \hat{a}$, which has expectation 
$\mathbb{E}(\hat{a}) = \frac{0.5 \cdot N}{0.5} = N$, as desired.  Now
consider case (2).  In this case, the probability of hauling out is 1.0 for
seals in the ice-covered area, so we have 
$\mathbb{E}(\hat{N}) = \frac{0.5 \cdot N}{1.0} = 0.5N$. That
is, we bias abundance because we are not accounting for seals that are away from
ice, and therefore have a zero chance of hauling out. The same logic holds for
any other covariate, such as distance from the ice edge, that has different
values in the surveyed and unsurveyed habitat of the seals (assuming distance
from ice edge is coded with opposite signs inside and outside the pack ice).
Lastly, our study was limited to the spring season when seal haul-out behaviors
are strongly influenced by pupping, nursing, breeding behavior, and molt and
these drivers are likely more influential than specific sea-ice concentration.
Crawford et al. [-@crawford2019] compared haul-out probability models for ringed
seals and found those that only included season (and not sea-ice concentration)
were the most parsimonious. For these reasons, we have elected not to use sea
ice concentration _as a predictor for haul-out probability_ in the present
study.
